Embedded Program Search in Agda

### 2020-04-25 paper/paper.tex
technical sections

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{The Name of the Title is Hope}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{TBD}
\email{tbd@colorado.edu}
\author{TBD}
\email{tbd@colorado.edu}
\affiliation{%
  \institution{University of Colorado Boulder}
}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  TBD
\end{abstract}

@@

\section{(Contribution 1)}

\section{(Contribution 2)}

@@

@@

>>>>

\usepackage{xargs}                      % Use more than one optional parameter in a new commands
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
\newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
% \newcommandx{\info}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}}
% \newcommandx{\improvement}[2][1=]{\todo[linecolor=Plum,backgroundcolor=Plum!25,bordercolor=Plum,#1]{#2}}
% \newcommandx{\thiswillnotshow}[2][1=]{\todo[disable,#1]{#2}}

\newcommand{\NC}{(\textbf{Needs Citation})}

\newcommand{\term}[1]{\textsf{#1}}
\newcommand{\type}[1]{\textsf{#1}}

\newcommand{\Carrier}{\term{Carrier}}
\newcommand{\Universe}{\mathcal{U}}
\newcommand{\Set}{\type{Set}}
\newcommand{\Type}{\type{Type}}
\newcommand{\Boolean}{\term{Boolean}}
\newcommand{\true}{\term{true}}
\newcommand{\false}{\term{false}}
\newcommand{\Number}{\term{Number}}
\newcommand{\Interp}[1]{\llbracket #1 \rrbracket}
\newcommand{\?}{\stackrel{?}{\approx}}
\newcommand{\search}{\term{search}}

\newcommand{\data}[2]{\textsf{data } #1 : #2 \textsf{ where}}
\newcommand{\record}[2]{\textsf{record } #1 : #2 \textsf{ where}}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Embedded Program Search in Agda}

%%
%% The "author" command and its associa

%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Jack Martin}
\email{John.P.Martin@colorado.edu}
\author{Michael Dresser}
\email{tbd@colorado.edu}
\affiliation{%
  \institution{University of Colorado Boulder}
}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
The problem of Program Search - looking for a function which satisfies a query -
is one of the more practical tools a developer can use when programming. Search
tools such as Hoogle provide an entrypoint for a language eco-system, but
unfortunately often don't accept query information other than types. Here we
present a search tool for searching over types and examples as well as a general
framework for doing program search over dependent types. We also provide a
type-theoretic interpretation of the program search problem as a special case of
the more general synthesis problem and discuss how this presents a tool for
furthering the solution to both problems.
\end{abstract}

@@

\section{Preliminaries}

\todo[inline]{Note for Benno and Evan: The citations are absent for now. We will
add these for the final paper but decided to leave them out for now until the
structure of the paper is fixed. The notes for ``Needs citation'' indicate
positions where we have a reference saved somewhere. It is not an indication
that we need to still find it. }

\subsection{Universes} One of the advantages of dependent type theories is their
ability to formalize their own type system and meta-theories within
themselves(\textbf{Needs citation}). These techniques typically use a structure
called a \textit{Universe}, which is defined as a type along with an
interpretation function which takes an element of that type and computes some
type in the type system. Formally:

$$
\begin{array}{l}
\record{\Universe}{\Set}\\
\quad  \begin{array}{rl}
         \Carrier &: \Set \\
         \Interp{\_} &: \Carrier \rightarrow \Set
       \end{array}
\end{array}
$$

An example of such a universe containing booleans, natural numbers, and
functions on either is given here:

$$
\begin{array}{cc}
\begin{array}{l}
\data{\Type}{\Set}\\
\quad \begin{array}{rl}
        \textsf{Number} :& \Type\\
        \textsf{Boolean} :& \Type\\
        \_\Rightarrow\_ :& \Type \rightarrow \Type \rightarrow \Type
      \end{array}
\end{array}
&
\begin{array}{l}
  \Interp{\_}_{\Type} : \Type \rightarrow \Set\\
  \Interp{\textsf{Number}}_{\Type}     = \mathbb{N}\\
  \Interp{\textsf{Boolean}}_{\Type}    = \mathbb{B}\\
  \Interp{t_1 \Rightarrow t_2}_{\Type} \ \, = \Interp{t_1} \rightarrow \Interp{t_2}
\end{array}
\end{array}
$$

Soon, we will define a universe of types for use in program search. We will
spend most of the rest of the paper defining a suitable $\Carrier$ type that
respects the various complexities that arise from the presence of dependent
types.

\subsection{Simulating First Class Modules}

We will need a coarse simulation of Agda's module structure to store entries for
search. This will be accomplished by collecting elements of a $\Carrier$ type
along with the terms they encode into a homogeneous list. The first element will
be the coded type, and the second element(which depends on the first), will be
the term of the type that is interpreted from the code:

$$
\textsf{Module} = \Sigma[ \tau \in \Carrier ] \Interp{ \tau }
$$

Carrying on with our example universe above, an example module containing some
common functions and elements:

$$
\begin{array}{rl}
  \textsf{example} =& (\Boolean\ ,\ \true)\\
                  ::& (\Number\ ,\ 5 )\\
                  ::& (\Boolean\ ,\ \false )\\
                  ::& (\Number \Rightarrow \Boolean\ ,\ (== 5) )\\
                  ::&[]
\end{array}
$$

\section{Searching over Universe Modules}

To support basic type search over these modules we will first have to discuss
what it means for a query type to ``match'' a type in a collection. This clearly
introduces the need for a notion of equality over types, which complicates
matters quickly. Which form of equality of types is appropriate for use in
program search? Fortunately, there has been some research done on this problem
and the current accepted equality is isomorphism of types\NC. This
will equate curried function types to their uncurried counterparts for
example(i.e. $(A \times B) \rightarrow C \approx A \rightarrow B \rightarrow
C$).

A problem with using isomorphism over dependent types is that it is a generally
undecidable problem\NC. The field is currently studying certain decidable
subsets of the problem thus we opt to maximize flexibility by parameterizing our
methods over a provided decidable type equality. The problem of type-directed
program search then just becomes the act of running the decision procedure for
this equality over the coded-types in our module. Thus, a searchable universe
can be defined:
\todo{define isEquiv}
$$
\begin{array}{l}
  \record{\Universe_S}{\Set}\\
  \quad \begin{array}{ll}
          U &: \Universe\\
         \_\approx\_ &: (a\ b\ : \Carrier\ U) \rightarrow \Set\\
         \approx-\textsf{isEquiv} &: \textsf{IsEquivalence } (\Carrier\ U)\ \_\approx\_\\
         \_\?\_ &: \forall (a\ b : \Carrier\ U) \rightarrow \textsf{Dec } (a \approx b)
        \end{array}
\end{array}
$$

Now we are ready to define an abstract search procedure for modules of a
decidable universe. The type of the search function is:

$$
\begin{array}{l}
  \search : (u : \Universe_S) \rightarrow (m : \textsf{Module}_u)\rightarrow (c : \Carrier\ (U\ u))
  \rightarrow \textsf{Maybe }\Interp{c}
\end{array}
$$

The procedure takes in a decidable universe $u$, a module of that universe $m$,
and a coded-type $c$. The output type of the function is possibly an inhabitant
of the interpretation of $c$ provided that there is in fact one in the module.

\section{A Universe for Searching over Dependent Types}

Thus far, the formalism we have given should allow any non-dependent program
to be searched. To add support for programs with type dependencies we will need
to slightly alter our definition of $\textsf{Module}$. The main idea is that the
module will need to be simultaneously defined with the interpretation function
so that types defined in a ``cons'' of a module can depend on the terms found in
earlier entries.

The definition of this universe is given in Figure \ref{fig1}.
\todo[inline]{Note for Benno and Evan: There is not any explanation or elaboration for this
  part of the formalism yet. There are also a few missing constructors. We were
  stuck on this for the better part of the last week and just figured out how to
formalize it last night and it isn't quite finished. In the final paper this
will be much, much more spelled out over 3-5 pages.}

\begin{figure}[t]
  $$
  \begin{array}{cc}
  \begin{array}{l}
    \data{\textsf{Module}}{\mathbb{N} \rightarrow \Set}\\
    \quad \begin{array}{ll}
            []        &: \textsf{Module } 0\\
            \_::\_,\_ &: \forall \{ n \}\\
                      &\rightarrow  (\Gamma : \textsf{Module } n)\\
                      &\rightarrow  (t : \textsf{TypeCode } n)\\
                      &\rightarrow  \Interp{ t }_\Gamma \\
                      &\rightarrow \textsf{Module (suc }  n)
          \end{array}
  \end{array}
                        &
  \begin{array}{l}
    \data{\textsf{TypeCode}}{\mathbb{N} \rightarrow \Set}\\
    \quad
    \begin{array}{ll}
      U &: \forall \{ n \} \rightarrow \textsf{TypeCode } n\\
      \# &: \forall \{ n\ m \} \rightarrow n < m \rightarrow \textsf{TypeCode } m\\
      \_\Rightarrow\_ &: \forall \{ n \}\ (t_1\ t_2 : \textsf{TypeCode } n)
                        \rightarrow \textsf{TypeCode } n
    \end{array}

  \end{array}
  \end{array}
  $$
  $$
  \begin{array}{l}
    \Interp{\_}\_ : \forall \{ n \} \rightarrow \textsf{TypeCode } n \rightarrow
    \textsf{Module } n \rightarrow \Set\\
    \Interp{ U }_\Gamma = \Set\\
    \Interp{ t_1 \Rightarrow t_2 }_\Gamma = \Interp{t_1}_\Gamma \rightarrow \Interp{t_2}_\Gamma\\
    \Interp{ \#\ \textsf{suc-leq-suc ( zero-leq-n)}}_{(\Gamma :: U ,
    \textsf{term})} = \textsf{term}\\
    \Interp{ \#\ \textsf{suc-leq-suc ( zero-leq-n)}}_{(\Gamma :: t ,
    \textsf{term})} = \langle\textsf{Error Case - Impossible under assumptions}\rangle\\
    \Interp{ \#\ \textsf{suc-leq-suc ( suc-leq-suc } n )}_{(\Gamma :: t ,
    \textsf{term})} = \Interp{ \#\ \textsf{suc-leq-suc } n }_\Gamma
  \end{array}
  $$
  \caption{The Universe for Search over Dependent Types}
  \label{fig1}
\end{figure}

@@

\todo[inline]{We were unsure what to include in this section because most of our
  project was correct by construction. Benno and Evan, we will touch base with
  you about your thoughts as to what to include here. We brainstormed a little
  bit but couldn't settle on anything that made too much sense. Hoping that the
  sections above will be suitable for this week's version of the draft. One idea
  we had would be to show off an example of why the embedded program search is
  so powerful/useful. This seems to be a somewhat empirical claim that we could
  spend some time fleshing out.}

@@

\todo[inline]{Edit: Make changes as Benno suggested}

\paragraph{Program Search} Program Search, or the search of functions in some
library, has led to such tools as Hoogle, and is the main category that this
work could be included under. In the past, none of these tools use any
functional properties or examples to help guide the search. Ours is the first
tool to provide this functionality.

\textit{draft note: make sure to look into the search programs for other languages.}

\paragraph{Program Synthesis} As was mentioned in Section \# above, program
search can be viewed as a special case of the more general program synthesis
problem. It seems hopeful that program search can be later used as a tool in
program synthesis by treating it as an atomic operation for larger synthesis
procedures.

\paragraph{Type Isomorphism} Isomorphisms over types acts as the equivalence
relation that we use for search in our approach. In reality our approach defines
a slightly more robust isomorphism relation over types \textit{and} examples.
This is slightly more powerful than just the types themselves. In future work we
will look at the theoretical basis for this.
